Time for a rethink: time sub-sampling methods in disparity-through-time analyses

### 2018-02-12 manuscript/time-slice.tex
Complete edits and reformat back to nice LaTeX layout

% Preamble
\documentclass[12pt,a4paper]{article}
\usepackage{enumerate} 	
\usepackage{setspace}						
\usepackage{authblk}	
\usepackage{graphicx} 	
\usepackage[nomarkers, nolists, figuresonly]{endfloat} 
\usepackage{pdflscape}	
\usepackage{mathtools}	
\usepackage[osf]{mathpazo} 
\usepackage{lineno} 
\usepackage{ms}    	
\usepackage{hyperref}
\usepackage[round]{natbib} 
\usepackage{setspace}

@@

Disparity-through-time analyses have two main analysis components: calculating disparity, and creating time sub-subsets of the data. 
Here we focus on the latter.
The nature of disparity (i.e. it is a diversity metric), means it cannot be calculated using a single individual, so some way of sub-sampling taxa is required.
Changes in disparity-through-time are generally investigated by calculating the disparity of taxa present during specific time intervals or time bins \citep[e.g][]{cisneros2010,prentice2011,Hughes20082013,hopkinsdecoupling2013,bentonmodels2014,bensonfaunal2014}.
These time bins are usually defined based on stratigraphy \citep[e.g.][]{cisneros2010,prentice2011,Hughes20082013,bentonmodels2014} but can also be arbitrarily chosen time bins of equal (or approximately equal) duration \citep{Butler2012,hopkinsdecoupling2013,bensonfaunal2014}.
However, this approach has several limitations.

@@

% overview figure
\begin{figure}[!htbp]
    \centering
    %\includegraphics[width=1\linewidth, height=1\textheight, keepaspectratio]{figures/Figure1-outlinesvg.pdf}
    \caption[Outline of the disparity-through-time pipeline]
    {Outline of the disparity-through-time pipeline. 1) We use ancestral character estimation to infer nodal character states; 2) we measure the pairwise Gower distance between the tip character states and nodal character states; 3) we ordinate the distance matrix using principal coordinates analysis (PCoA/PCO); 4) we time sub-sample the PCoA matrix using time bins defined by stratigraphic periods, equally sized time bins and time-slices (using six methods to estimate ordination scores for branches); and finally 5) we measure disparity-through-time for each of these methods.}
    \label{fig:overview}
\end{figure}

@@

% Dataset details table
%\begin{landscape}
%\input{tables/table_datasets.tex} 
  % Label in input or referencing doesn't work
%\end{landscape}

@@

\paragraph{Estimating ancestral character states.}
\label{ace}
For each dataset we estimated the ancestral character states at each node using the \texttt{AncStatesEstMatrix} function from the \texttt{Claddis} \texttt{R} package \citep{Claddis,R}.
This function uses the re-rooting method \citep{Yang01121995,Garland2000} to get Maximum Likelihood estimates of the ancestral states for each character at every node in the phylogeny \citep[based on the \texttt{rerootingMethod} function in \texttt{phytools};][]{phytools}.
Inapplicable and missing characters for any taxon were treated as ambiguous characters (i.e. any possible observed state for the character).
To prevent poor ancestral state estimations from biasing our results, especially when a lot of error is associated with the estimations, we only included ancestral state estimations with a scaled Likelihood $\geq$ $0.95$.
Ancestral state estimations with scaled Likelihoods below this threshold were recoded as missing data (``?'').
This allowed our results to be less dependent on the quality (or the absence thereof) of the ancestral state estimations methods, especially in parts of the datasets where data were sparse. 
This approach is similar to \cite{brusatte2011phylogenetic} but uses model based estimations (rather than parsimony) allowing to control for ambiguous nodes (i.e. poorly estimated).

\paragraph{Building morphospaces.} 
To explore disparity-through-time in our datasets, we used a morphospace approach \citep[e.g.][]{Foote01071994,Foote29111996,Wesley-Hunt2005,Brusatte12092008,friedmanexplosive2010,toljagictriassic-jurassic2013,Hughes20082013}.
Morphospaces can be obtained from any multidimensional morphological data set but can differ in the data used (e.g. discrete or continuous), and whether they include phylogenetic data or not.
Although empirical morphospaces from discrete or continuous data have been shown to have similar properties \citep{foth2012different,hetherington2015cladistic}, our morphospaces are based on discrete morphological data (originally collected for phylogenetic analysis; \textit{c.f.} geometric morphometric data) and includes some phylogenetic information (see above).
Mathematically, our morphospaces are $n$ dimensional objects that summarise the distances between discrete morphological characters of the taxa present and their ancestors.

\paragraph{Constructing distance matrices.}
To estimate the morphospaces for each of our datasets we first constructed pairwise distance matrices of length $n$, where $n$ is the total number of tips and nodes in the dataset.
We calculated the $n$$\times$$n$ distances using the Gower distance \citep{Gower71}, i.e. the number of characters mismatched characters over the number of shared characters.
This allows us to correct for distances between two taxa that share many characters and could be closer to each other than to taxa with fewer characters in common (i.e. because some pairs of taxa share more characters in common than others, they are more likely to be similar).
For discrete morphological matrices, using this corrected distance is preferable to the raw Euclidean distance because of its ability to deal with discrete or/and ordinated characters as well as with missing data \citep{anderson2012using}.
However, the Gower distance cannot calculate distances when taxa have no overlapping data.
Therefore, we used the \texttt{TrimMorphDistMatrix} function from the \texttt{Claddis} \texttt{R} package to remove pairs of taxa with no cladistic characters in common.
This led to us removing nine taxa from the \cite{bapst2016topology} dataset, and 19 from the \cite{brusatte2014gradual} dataset, but none from the other two datasets (see Supporting Information Appendix S1 for details of which species).

@@

To compare the time binning and time-slicing approaches we applied the methods as follows (see Figure \ref{figure:bins}).

\begin{figure}[!htbp]
    \centering
    %\includegraphics[width=1\linewidth, height=1\textheight, keepaspectratio]{figures/Figure2-binning.pdf}
    \caption[Outline of the three time sub-sampling methods]
    {Outline of the three time sub-sampling methods. 
    Stratigraphy: time sub-samples are defined by stratigraphic periods.
    Here there are five stratigraphic periods in the 20 My (million years) time frame of interest, i.e. five bins/slices with variable sizes/intervals.
    Duration: time sub-samples are defined based on the mean duration of stratigraphic periods in the time frame of interest. 
    Here, the mean duration of stratigraphic periods is 5 My, so there are four bins/slices of 5 My duration (or four slices with 5 My intervals between them) in the 20 My time frame of interest. 
    Number: time sub-samples are defined based on the number of stratigraphic periods in the time frame of interest. 
    Here, there are five stratigraphic periods, so there are five bins/slices of 4 My duration (or five slices with 4 My intervals between them) in the 20 My time frame of interest.}
    \label{fig:bins}
\end{figure}

\begin{enumerate}
  \item Stratigraphy: time sub-samples defined by stratigraphic periods. 
  \begin{enumerate}[(i)]
    \item Time bins (unequal). 
    We calculated disparity for the taxa in each stratigraphic period (stage or epoch). 
    To reduce the influence of outliers on our disparity estimates, we bootstrapped each disparity measurement for each time bin by randomly resampling with replacement a new sub-sample of taxa from the observed taxa in the bin 100 times.
    We then calculated the median disparity value for each time bin along with the 50\% and 95\% confidence intervals.
    \item Time-slices (non-equidistant).
    We calculated disparity using our time-slicing approach with slices occurring at the midpoint of each stratigraphic period (stage or epoch), and using all six time-slicing methods (acctran, deltran, random, proximity, equal splits and gradual splits).
    To reduce the influence of outliers on our disparity estimates, we bootstrapped each disparity measurement as described above for the stratigraphic time bins.
    \end{enumerate}

  \item Duration: time sub-samples defined by the duration of stratigraphic periods. 
  \begin{enumerate}[(i)]
    \item Time bins (equal). 
    We calculated disparity for the taxa in each time bin where time bin size was defined by the mean \textit{duration} of the stratigraphic period (stage or epoch), and bootstrapped the disparity values as described above.
    \item Time-slices (equidistant).
    We calculated disparity using our time-slicing approach where the interval between slices, was defined by the mean \textit{duration} of the stratigraphic period (stage or epoch).
    We used the six time-slicing methods and bootstrapped as described above.
  \end{enumerate}

  \item Number: time sub-samples defined by the number of stratigraphic periods. 
  \begin{enumerate}[(i)]
    \item Time bins (equal). 
    We calculated disparity for the taxa in each time bin where the number of time bins was defined by the \textit{number} of stratigraphic periods (ages or epochs) in the time frame of interest, and bootstrapped the disparity values as described above.
    \item Time-slices (equidistant).
    We calculated disparity using our time-slicing approach where the number of slices, was defined by the \textit{number} of stratigraphic periods (ages or epochs) in the time frame of interest.
    We used the six time-slicing methods and bootstrapped as described above.
  \end{enumerate}

@@

\begin{figure}[!htbp]
    \centering
    %\includegraphics[width=1\linewidth, height=1\textheight, keepaspectratio]{figures/Figure2-fig-dtt-epoch-revision.pdf}
    \caption[Relative disparity-through-time.]
    {Relative disparity-through-time. Median bootstrapped disparities were calculated using time binning and time-slicing approaches. 
    Green points represent time binning methods, purple points are time-slices with a punctuated model of evolution (`proximity' method), and blue points are time-slices with a gradual model of evolution (`gradual splits' method).
    Relative disparities (median bootstrapped disparity divided by the maximum median bootstrapped disparity for a dataset and analysis method) are presented so they can be compared across datasets/methods. 
    Stratigraphy uses unequal time bins or non-equidistant time-slices, where the width of the bin, or the interval between slices, is equivalent to stratigraphic epochs. 
    Duration uses equal time bins or equidistant time-slices, where the width of the bin, or the interval between slices, is the average duration of stratigraphic epochs in the time frame of the dataset. 
    Number uses equal time bins or equidistant time-slices, where the number of bins, or the number of slices, is the average number of stratigraphic epochs in the time frame of the dataset. 
    In all cases, time bin disparities are plotted at the midpoint of the bin, and error bars represent the 95\% confidence intervals around the bootstrapped median disparity.
    The four dataset names are on the first plot for each dataset (see Table \ref{table:datasets} for details).
    Results for stratigraphic stages, and for other time-slicing methods, are in the Supporting Information Appendix S2: Figures A1-A2.}
    \label{figure:dtt1}
\end{figure}

@@

\paragraph{Systematic differences in disparity-through-time.} 
There is no overall significant systematic difference among the disparities calculated using time bins and those calculated using the time-slicing methods (Table \ref{table:wilcox}, Supporting Information Appendix S2: Table A1).
Instead, the differences depend on the dataset and method in question.
For example, the \cite{brusatte2014gradual}, \cite{bapst2016topology} and \cite{wright2017bayesian} datasets, show significant differences when using bins versus time-slices defined by stratigraphy, but the \cite{beckancient2014} dataset appears robust to these different approaches.
Likewise, the \cite{beckancient2014}, \cite{brusatte2014gradual} and \cite{bapst2016topology} datasets have different disparities when the number of bins or slices is the average number of stratigraphic periods, but this is not seen in the \cite{wright2017bayesian} dataset.
Note that for epochs, we find fewer significant differences simply because the smaller numbers of bins and slices being compared means we have low power to detect a significant difference.

% Wilcoxon results table
%\input{tables/table_wilcox.tex}

\paragraph{Disparity peaks.}
In the \cite{beckancient2014} and \cite{bapst2016topology} datasets, disparity peaks occur much at much older ages when time-slicing rather than time binning approaches are used (Fig. \ref{figure:peak1}; Supporting Information Appendix S2: Figs A3-A4).
This is also true for stratigraphic time bins in the \cite{wright2017bayesian} dataset, although when using equal time bins the peaks are later than the time-slicing methods, or very similar (Fig. \ref{figure:peak1}; Supporting Information Appendix S2: Figs A3-A4).
Across the three time binning methods, the \cite{brusatte2014gradual} dataset has similar disparity peaks whichever method is used, the \cite{wright2017bayesian} dataset only had variation in peaks when using unequal time bins (stratigraphy), whereas in the \cite{bapst2016topology} and \cite{beckancient2014} datasets, stratigraphic (unequal) versus equally sized time bins make a large difference to where the disparity peak occurs (Fig. \ref{figure:peak1}; Supporting Information Appendix S2: Figs A3-A4).
Additionally, there seems to be small discrepancies within the time-slicing methods (gradual.split \textit{vs.} proximity) except in the \cite{beckancient2014} dataset where the gradual split model recovered disparity peaks at younger ages than the proximity model (Fig. \ref{figure:peak1}; Supporting Information Appendix S2: Figs A3-A4)

% Plots showing the differences
\begin{figure}[!htbp]
    \centering
    %\includegraphics[width=1\linewidth, height=1\textheight, keepaspectratio]{figures/Figure3-fig-peaks-epoch-revision.pdf}
    \caption[Timing of peak disparity for four example datasets.]
    {Timing of peak disparity.
    Median bootstrapped disparities were calculated using time binning and time-slicing approaches. 
    Green points represent time binning methods, blue points are time-slices with a punctuated model of evolution (`proximity' method), and purple points are time-slices with a gradual model of evolution (`gradual splits' method). 
    Stratigraphy uses unequal time bins or non-equidistant time-slices, where the width of the bin, or the interval between slices, is equivalent to stratigraphic epochs. 
    Duration uses equal time bins or equidistant time-slices, where the width of the bin, or the interval between slices, is the average duration of stratigraphic epochs in the time frame of the dataset. 
    Number uses equal time bins or equidistant time-slices, where the number of bins, or the number of slices, is the average number of stratigraphic epochs in the time frame of the dataset. 
    For time bins the points indicate the maximum and minimum ages of the time bin within which peak disparities appeared.
    The four dataset names are on the first plot for each dataset (see Table \ref{table:datasets} for details).
    Results for stratigraphic stages, and for other time-slicing methods, are in the Supporting Information Appendix S2: Figures A3-A4.}
    \label{figure:peak1}
\end{figure}

@@

% Big heat map table thing
\begin{figure}[!htbp]
    \centering
    %\includegraphics[width=1\linewidth, height=1\textheight, keepaspectratio]{figures/Figure4-fig-extinction-effect.pdf}
    \caption[Extinction example datasets.]
    {Effects of mass extinction events on disparity.
    Pink cells and blue cells indicate respectively a significant or non significant change in disparity before and after the mass extinction event (Cretaceous-Paleogene 66 MYA; \citealt{beckancient2014}; Ordovician-Silurian 455-430 MYA; \citealt{wright2017bayesian}).
    \textit{e:1}, \textit{e:2}, and \textit{e:3} denote whether the comparison was between the time bin or time-slice immediately after the mass extinction (\textit{e:1}), or the second (\textit{e:2}) or third (\textit{e:3}) bin/slice after the mass extinction to account for any lag effect.
    The top seven rows use stratigraphic stages and the bottom seven rows use stratigraphic epochs.
    Labels on the left hand side indicate whether time bins (``bins'') were used, or which of the six time-slicing methods was used.}
    \label{figure:extinction1}
\end{figure}

\section{Discussion}
Disparity-through-time analyses are influenced by the choice of time sub-sampling method used to divide the taxa. 
While differences in the relative disparities calculated among time sub-sampling methods may not be of much biological importance, these changes can have important implications for the conclusions of downstream analyses. 
For example, using stratigraphic epochs as our reference time period, there are 21 potential methods for time sub-sampling our data (splitting by stratigraphy, number and duration, see methods, and using time bins or one of six time-slicing methods). 
Of these 21 methods, in 16 (76\%) we show that placental mammals \citep{beckancient2014} significantly increased in disparity in the time bin/slice immediately after the K-Pg mass extinction event, and in 20 (95\%) we show that crinoids \citep{wright2017bayesian} significantly decreased in disparity in the time bin/slice immediately after the O-S mass extinction event.
Given the high congruence (76\% and 95\%) of these results, one could argue that time-sub-sampling methods are not important. 
However, if we had chosen to investigate crinoid disparity only using time bins and splitting these so the number of time bins was equal to the number of epochs (number), we would have concluded that the O-S extinction had no effect on crinoid disparity. 
Likewise, the timing of peak disparity differs among methods. 
This is particularly evident when comparing stratigraphic time bins to time slicing methods, where for most of our datasets we see a much later time to peak disparity. 
This could have major implications for our understanding of how morphological diversity changes through time, for example in response to climate effects.
These results highlight the sensitivity of disparity-through-time analyses to the choice of time sub-sampling method. 
Fortunately this issue is easy to solve; either disparity-through-time analyses should use, and report results from, multiple time sub-sampling methods (as demonstrated here), or great care should be taken in determining the appropriate time sub-samples to answer the question of interest.

@@

Of course the method also has limitations.
The main one of these is a practical one; it requires a time-calibrated phylogeny which are not available for all palaeontological datasets.
Furthermore, like most phylogeny based methods, time-slicing depends on ancestral state estimations.
Care should be taken in interpreting these, as they are highly dependent on the data and models used for the estimations \citep{Slateretal2012,doi:10.1080/10635150801910451}.
The difference between the time-binning and time-slicing results could also simply be due to the nature of the fossil record.
Rates of sedimentation vary in time and space influencing the groups found within the rock record and their temporal distribution. 
In this case, different beds could represent different ``packages'' of fauna through time separated by gaps, resulting in natural ``bins'' rather than slices of the data.
Slicing through such strata will yield similar results no matter where in time the slice occurs.
It is important to note however, that the time slicing method also includes ancestral estimations (either through the nodes or the branches) that are by definition not available in the fossil record and thus are not influenced by its nature.
Additionally, this effect is likely to be most obvious in vertebrates or groups where the fossil is ``patchy'' but less problematic for groups with a more continuous one like Foraminifera.
Finally, \cite{Hunt21042015} found that time series are best characterized by gradual directional changes (biased random walks).
In fact, homogeneous directional changes are more likely to be supported than heterogeneous ones (e.g. punctuated changes) in longer duration series with few samples per series.
In our implementation of time-slicing, the models are not selected based on any fit value (e.g. AIC) but merely on researcher's assumptions.
We thus suggest that both types of models (punctuated and gradual) are tested during analysis unless there could be strong independent support for one or the other.

@@

\newpage
\section{Figure legends}

>>>>

% Preamble
\documentclass[12pt,a4paper]{article}
\usepackage{enumerate} 	
\usepackage{setspace}						
\usepackage{authblk}	
\usepackage{graphicx} 	
%\usepackage[nomarkers, nolists, figuresonly]{endfloat} 
\usepackage{pdflscape}	
\usepackage{mathtools}	
\usepackage[osf]{mathpazo} 
\usepackage{lineno} 
\usepackage{ms}    	
\usepackage{hyperref}
\usepackage[round]{natbib} 
\usepackage{setspace}

@@

Disparity-through-time analyses have two main analysis components: calculating disparity, and creating time sub-subsets of the data. 
Here we focus on the latter.
The nature of disparity (i.e. it is a diversity metric), means it cannot be calculated using a single individual, so some way of sub-sampling taxa is required.
Changes in disparity-through-time are generally investigated by calculating the disparity of taxa present during specific time intervals or time bins \citep[e.g.][]{cisneros2010,prentice2011,Hughes20082013,hopkinsdecoupling2013,bentonmodels2014,bensonfaunal2014}.
These time bins are usually defined based on stratigraphy \citep[e.g.][]{cisneros2010,prentice2011,Hughes20082013,bentonmodels2014} but can also be arbitrarily chosen time bins of equal (or approximately equal) duration \citep{Butler2012,hopkinsdecoupling2013,bensonfaunal2014}.
However, this approach has several limitations.

@@

% overview figure
\begin{figure}[!htbp]
    \centering
    \includegraphics[width=1\linewidth, height=1\textheight, keepaspectratio]{figures/Figure1-outlinesvg.pdf}
    \caption[Outline of the disparity-through-time pipeline]
    {Outline of the disparity-through-time pipeline. 1) We use ancestral character estimation to infer nodal character states; 2) we measure the pairwise Gower distance between the tip character states and nodal character states; 3) we ordinate the distance matrix using principal coordinates analysis (PCoA/PCO); 4) we time sub-sample the PCoA matrix using time bins defined by stratigraphic periods, equally sized time bins and time-slices (using six methods to estimate ordination scores for branches); and finally 5) we measure disparity-through-time for each of these methods.}
    \label{fig:overview}
\end{figure}

@@

% Dataset details table
\begin{landscape}
\input{tables/table_datasets.tex} 
  % Label in input or referencing doesn't work
\end{landscape}

@@

\paragraph{Estimating ancestral character states.}
\label{ace}
For each dataset we estimated the ancestral character states at each node using the \texttt{AncStatesEstMatrix} function from the \texttt{Claddis} \texttt{R} package \citep{Claddis,R}.
This function uses the re-rooting method \citep{Yang01121995,Garland2000} to get Maximum Likelihood estimates of the ancestral states for each character at every node in the phylogeny \citep[based on the \texttt{rerootingMethod} function in \texttt{phytools};][]{phytools}.
Inapplicable and missing characters for any taxon were treated as ambiguous characters (i.e. any possible observed state for the character).
To prevent poor ancestral state estimations from biasing our results, especially when a lot of error is associated with the estimations, we only included ancestral state estimations with a scaled Likelihood $\geq$ $0.95$.
Ancestral state estimations with scaled Likelihoods below this threshold were recoded as missing data (``?'').
This allowed our results to be less dependent on the quality (or the absence thereof) of the ancestral state estimations methods, especially in parts of the datasets where data were sparse. 
This approach is similar to \cite{brusatte2011phylogenetic} but uses model based estimations (rather than parsimony) allowing us to control for ambiguous (i.e. poorly estimated) nodes.

\paragraph{Building morphospaces.} 
To explore disparity-through-time in our datasets, we used a morphospace approach \citep[e.g.][]{Foote01071994,Foote29111996,Wesley-Hunt2005,Brusatte12092008,friedmanexplosive2010,toljagictriassic-jurassic2013,Hughes20082013}.
Morphospaces can be obtained from any multidimensional morphological data set but can differ in the data used (e.g. discrete or continuous), and whether they include phylogenetic data or not.
Although empirical morphospaces from discrete or continuous data have been shown to have similar properties \citep{foth2012different,hetherington2015cladistic}, our morphospaces are based on discrete morphological data (originally collected for phylogenetic analysis; \textit{c.f.} geometric morphometric data) and include some phylogenetic information (see above).
Mathematically, our morphospaces are $n$ dimensional objects that summarise the distances between discrete morphological characters of the taxa present and their ancestors.

\paragraph{Constructing distance matrices.}
To estimate the morphospaces for each of our datasets we first constructed pairwise distance matrices of length $n$, where $n$ is the total number of tips and nodes in the dataset.
We calculated the $n$$\times$$n$ distances using the Gower distance \citep{Gower71}, i.e. the number of mismatched characters over the number of shared characters.
This allows us to correct for distances between two taxa that share many characters and could be closer to each other than to taxa with fewer characters in common (i.e. because some pairs of taxa share more characters in common than others, they are more likely to be similar).
For discrete morphological matrices, using this corrected distance is preferable to the raw Euclidean distance because of its ability to deal with discrete or/and ordinated characters as well as with missing data \citep{anderson2012using}.
However, the Gower distance cannot calculate distances when taxa have no overlapping data.
Therefore, we used the \texttt{TrimMorphDistMatrix} function from the \texttt{Claddis} \texttt{R} package to remove pairs of taxa with no cladistic characters in common.
This led to us removing nine taxa from the \cite{bapst2016topology} dataset, and 19 from the \cite{brusatte2014gradual} dataset, but none from the other two datasets (see Supporting Information Appendix S1 for details of which species).

@@

To compare the time binning and time-slicing approaches we applied the methods as follows (see Fig. \ref{figure:bins}).

\begin{figure}[!htbp]
    \centering
    \includegraphics[width=1\linewidth, height=1\textheight, keepaspectratio]{figures/Figure2-binning.pdf}
    \caption[Outline of the three time sub-sampling methods]
    {Outline of the three time sub-sampling methods. 
    Stratigraphy: time sub-samples are defined by stratigraphic periods.
    Here there are five stratigraphic periods in the 20 My (million years) time frame of interest, i.e. five bins/slices with variable sizes/intervals.
    Duration: time sub-samples are defined based on the mean duration of stratigraphic periods in the time frame of interest. 
    Here, the mean duration of stratigraphic periods is 5 My, so there are four bins/slices of 5 My duration (or four slices with 5 My intervals between them) in the 20 My time frame of interest. 
    Number: time sub-samples are defined based on the number of stratigraphic periods in the time frame of interest. 
    Here, there are five stratigraphic periods, so there are five bins/slices of 4 My duration (or five slices with 4 My intervals between them) in the 20 My time frame of interest.}
    \label{fig:bins}
\end{figure}

\begin{enumerate}
  \item Stratigraphy: time sub-samples defined by stratigraphic periods (Fig. \ref{fig:bins}). 
  \begin{enumerate}[(i)]
    \item Time bins (unequal). 
    We calculated disparity for the taxa in each stratigraphic period (stage or epoch). 
    To reduce the influence of outliers on our disparity estimates, we bootstrapped each disparity measurement for each time bin by randomly resampling with replacement a new sub-sample of taxa from the observed taxa in the bin 100 times.
    We then calculated the median disparity value for each time bin along with the 50\% and 95\% confidence intervals.
    \item Time-slices (non-equidistant).
    We calculated disparity using our time-slicing approach with slices occurring at the midpoint of each stratigraphic period (stage or epoch), and using all six time-slicing methods (acctran, deltran, random, proximity, equal splits and gradual splits).
    To reduce the influence of outliers on our disparity estimates, we bootstrapped each disparity measurement as described above for the stratigraphic time bins.
    \end{enumerate}

  \item Duration: time sub-samples defined by the duration of stratigraphic periods (Fig. \ref{fig:bins}). 
  \begin{enumerate}[(i)]
    \item Time bins (equal). 
    We calculated disparity for the taxa in each time bin where time bin size was defined by the mean \textit{duration} of the stratigraphic period (stage or epoch), and bootstrapped the disparity values as described above.
    \item Time-slices (equidistant).
    We calculated disparity using our time-slicing approach where the interval between slices, was defined by the mean \textit{duration} of the stratigraphic period (stage or epoch).
    We used the six time-slicing methods and bootstrapped as described above.
  \end{enumerate}

  \item Number: time sub-samples defined by the number of stratigraphic periods (Fig. \ref{fig:bins}). 
  \begin{enumerate}[(i)]
    \item Time bins (equal). 
    We calculated disparity for the taxa in each time bin where the number of time bins was defined by the \textit{number} of stratigraphic periods (ages or epochs) in the time frame of interest, and bootstrapped the disparity values as described above.
    \item Time-slices (equidistant).
    We calculated disparity using our time-slicing approach where the number of slices, was defined by the \textit{number} of stratigraphic periods (ages or epochs) in the time frame of interest.
    We used the six time-slicing methods and bootstrapped as described above.
  \end{enumerate}

@@

\begin{figure}[!htbp]
    \centering
    \includegraphics[width=1\linewidth, height=1\textheight, keepaspectratio]{figures/Figure3-fig-dtt-epoch-revision.pdf}
    \caption[Relative disparity-through-time.]
    {Relative disparity-through-time. Median bootstrapped disparities were calculated using time binning and time-slicing approaches. 
    Green points represent time binning methods, purple points are time-slices with a punctuated model of evolution (`proximity' method), and blue points are time-slices with a gradual model of evolution (`gradual splits' method).
    Relative disparities (median bootstrapped disparity divided by the maximum median bootstrapped disparity for a dataset and analysis method) are presented so they can be compared across datasets/methods. 
    Stratigraphy uses unequal time bins or non-equidistant time-slices, where the width of the bin, or the interval between slices, is equivalent to stratigraphic epochs. 
    Duration uses equal time bins or equidistant time-slices, where the width of the bin, or the interval between slices, is the mean duration of stratigraphic epochs in the time frame of the dataset. 
    Number uses equal time bins or equidistant time-slices, where the number of bins, or the number of slices, is the mean number of stratigraphic epochs in the time frame of the dataset. 
    In all cases, time bin disparities are plotted at the midpoint of the bin, and error bars represent the 95\% confidence intervals around the bootstrapped median disparity.
    The four dataset names are on the first plot for each dataset (see Table \ref{table:datasets} for details).
    Results for stratigraphic stages, and for other time-slicing methods, are in the Supporting Information Appendix S2: Figures A1-A2.}
    \label{figure:dtt1}
\end{figure}

@@

\paragraph{Systematic differences in disparity-through-time.} 
There is no overall significant systematic difference among the disparities calculated using time bins and those calculated using the time-slicing methods (Table \ref{table:wilcox}, Supporting Information Appendix S2: Table A1).
Instead, the differences depend on the dataset and method in question.
For example, the \cite{brusatte2014gradual}, \cite{bapst2016topology} and \cite{wright2017bayesian} datasets, show significant differences when using bins versus time-slices defined by stratigraphy, but the \cite{beckancient2014} dataset appears robust to these different approaches.
Likewise, the \cite{beckancient2014}, \cite{brusatte2014gradual} and \cite{bapst2016topology} datasets have different disparities when the number of bins or slices is the mean number of stratigraphic periods, but this is not seen in the \cite{wright2017bayesian} dataset.
Note that for epochs, we find fewer significant differences simply because the smaller numbers of bins and slices being compared means we have low power to detect a significant difference.

% Wilcoxon results table
\input{tables/table_wilcox.tex}

\paragraph{Disparity peaks.}
In the \cite{beckancient2014} and \cite{bapst2016topology} datasets, disparity peaks occur much at much older ages when time-slicing rather than time binning approaches are used (Fig. \ref{figure:peak1}; Supporting Information Appendix S2: Figs A3-A4).
This is also true for stratigraphic time bins in the \cite{wright2017bayesian} dataset, although when using equal time bins the peaks are later than the time-slicing methods, or very similar (Fig. \ref{figure:peak1}; Supporting Information Appendix S2: Figs A3-A4).
Across the three time binning methods, the \cite{brusatte2014gradual} dataset has similar disparity peaks whichever method is used, the \cite{wright2017bayesian} dataset only had variation in peaks when using unequal time bins (stratigraphy), whereas in the \cite{bapst2016topology} and \cite{beckancient2014} datasets, stratigraphic (unequal) versus equally sized time bins make a large difference to where the disparity peak occurs (Fig. \ref{figure:peak1}; Supporting Information Appendix S2: Figs A3-A4).
Additionally, there seem to be small discrepancies within the time-slicing methods (gradual.split \textit{vs.} proximity) except in the \cite{beckancient2014} dataset where the gradual split model recovered disparity peaks at younger ages than the proximity model (Fig. \ref{figure:peak1}; Supporting Information Appendix S2: Figs A3-A4)

% Plots showing the differences
\begin{figure}[!htbp]
    \centering
    \includegraphics[width=1\linewidth, height=1\textheight, keepaspectratio]{figures/Figure4-fig-peaks-epoch-revision.pdf}
    \caption[Timing of peak disparity for four example datasets.]
    {Timing of peak disparity.
    Median bootstrapped disparities were calculated using time binning and time-slicing approaches. 
    Green points represent time binning methods, blue points are time-slices with a punctuated model of evolution (`proximity' method), and purple points are time-slices with a gradual model of evolution (`gradual splits' method). 
    Stratigraphy uses unequal time bins or non-equidistant time-slices, where the width of the bin, or the interval between slices, is equivalent to stratigraphic epochs. 
    Duration uses equal time bins or equidistant time-slices, where the width of the bin, or the interval between slices, is the mean duration of stratigraphic epochs in the time frame of the dataset. 
    Number uses equal time bins or equidistant time-slices, where the number of bins, or the number of slices, is the mean number of stratigraphic epochs in the time frame of the dataset. 
    For time bins the points indicate the maximum and minimum ages of the time bin within which peak disparities appeared.
    The four dataset names are on the first plot for each dataset (see Table \ref{table:datasets} for details).
    Results for stratigraphic stages, and for other time-slicing methods, are in the Supporting Information Appendix S2: Figures A3-A4.}
    \label{figure:peak1}
\end{figure}

@@

% Big heat map table thing
\begin{figure}[!htbp]
    \centering
    \includegraphics[width=1\linewidth, height=1\textheight, keepaspectratio]{figures/Figure5-fig-extinction-effect.pdf}
    \caption[Extinction example datasets.]
    {Effects of mass extinction events on disparity.
    Pink cells and blue cells indicate respectively a significant or non significant change in disparity before and after the mass extinction event (Cretaceous-Paleogene 66 MYA; \citealt{beckancient2014}; Ordovician-Silurian 455-430 MYA; \citealt{wright2017bayesian}).
    \textit{e:1}, \textit{e:2}, and \textit{e:3} denote whether the comparison was between the time bin or time-slice immediately after the mass extinction (\textit{e:1}), or the second (\textit{e:2}) or third (\textit{e:3}) bin/slice after the mass extinction to account for any lag effect.
    The top seven rows use stratigraphic stages and the bottom seven rows use stratigraphic epochs.
    Labels on the left hand side indicate whether time bins (``bins'') were used, or which of the six time-slicing methods was used.}
    \label{figure:extinction1}
\end{figure}

\section{Discussion}
Disparity-through-time analyses are influenced by the choice of time sub-sampling method used to divide the taxa. 
While differences in the relative disparities calculated among time sub-sampling methods may not be of much biological importance, these changes can have important implications for the conclusions of downstream analyses. 
For example, using stratigraphic epochs as our reference time period, there are 21 potential methods for time sub-sampling our data (splitting by stratigraphy, number and duration, see methods, and using time bins or one of six time-slicing methods). 
Of these 21 methods, in 16 (76\%) we show that placental mammals \citep{beckancient2014} significantly increased in disparity in the time bin/slice immediately after the K-Pg mass extinction event, and in 20 (95\%) we show that crinoids \citep{wright2017bayesian} significantly decreased in disparity in the time bin/slice immediately after the O-S mass extinction event.
Given the high congruence (76\% and 95\%) of these results, one could argue that time-sub-sampling methods are not important. 
However, if we had chosen to investigate crinoid disparity only using time bins and splitting these so the number of time bins was equal to the number of epochs (number), we would have concluded that the O-S extinction had no effect on crinoid disparity. 
Likewise, the timing of peak disparity differs among methods. 
This is particularly evident when comparing stratigraphic time bins to time slicing methods, where for most of our datasets we see a much later time to peak disparity. 
This could have major implications for our understanding of how morphological diversity changes through time, for example in response to climate.
These results highlight the sensitivity of disparity-through-time analyses to the choice of time sub-sampling method. 
Fortunately this issue is easy to solve; either disparity-through-time analyses should use, and report results from, multiple time sub-sampling methods (as demonstrated here), or great care should be taken in determining the appropriate time sub-samples to answer the question of interest.

@@

Of course the method also has limitations.
The main one of these is a practical one; it requires a time-calibrated phylogeny and these are not available for all palaeontological datasets.
Furthermore, like most phylogeny based methods, time-slicing depends on ancestral state estimations.
Care should be taken in interpreting these, as they are highly dependent on the data and models used for the estimations \citep{Slateretal2012,doi:10.1080/10635150801910451}.
The difference between the time-binning and time-slicing results could also simply be due to the nature of the fossil record.
Rates of sedimentation vary in time and space influencing the groups found within the rock record and their temporal distribution. 
In this case, different beds could represent different ``packages'' of fauna through time separated by gaps, resulting in natural ``bins'' rather than slices of the data.
Slicing through such strata will yield similar results no matter where in time the slice occurs.
It is important to note however, that the time slicing method also includes ancestral estimations (either through the nodes or the branches) that are by definition not available in the fossil record and thus are not influenced by its nature.
Additionally, this effect is likely to be most obvious in groups where the fossil record is ``patchy'', e.g. vertebrates, but less problematic for groups with a more continuous record like Foraminifera.
Finally, \cite{Hunt21042015} found that time series are best characterized by gradual directional changes (biased random walks).
In fact, homogeneous directional changes are more likely to be supported than heterogeneous ones (e.g. punctuated changes) in longer duration series with few samples in each series.
In our implementation of time-slicing, the models are not selected based on any model fit criterion (e.g. AIC) but merely on researcher assumptions.
We thus suggest that both types of models (punctuated and gradual) are tested during analysis, unless there is strong independent support for one or the other.

@@
